Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-10-23 paper-new/intro.tex
more paper revisions

\eric{This paragraph could be deleted}
%% In order for this to work as expected, gradually-typed languages should allow for
%% different parts of the codebase to be in different places along the spectrum from
%% dynamic to static, and allow for those different parts to interact with one another.
%% Moreover, gradually-typed languages should support the smooth migration from
%% dynamic typing to static typing, in that the programmer can initially leave off the
%% typing annotations and provide them later without altering the meaning of the program.
%% % Sound gradual typing
%% Furthermore, the parts of the program that are written in a dynamic
%% style should soundly interoperate with the parts that are written in a
%% static style.  That is, the interaction between the static and dynamic
%% components of the codebase should preserve, to the extent possible,
%% the guarantees made by the static types.  In particular, while
%% statically-typed code can error at runtime in a gradually-typed
%% language, such an error can always be traced back to a
%% dynamically-typed term that violated the typing contract imposed by
%% statically typed code. Further, static type assertions are sound in
%% the static portion, and should enable type-based reasoning and
%% optimization.

@@

% Graduality property
One of the fundamental theorems for gradually typed languages is
\emph{graduality}, also known as the \emph{dynamic gradual guarantee},
originally defined by Siek, Vitousek, Cimini, and Boyland
\cite{siek_et_al:LIPIcs:2015:5031, new-ahmed2018}.
%
Informally, graduality says that migrating code from dynamical to
statical typing should only allow for the introduction of static or
dynamic type errors, and not otherwise change the behavior of the
program.
%
This is a way to capture programmer intuition that increasing type
precision corresponds to a generalized form of runtime assertions in
that there are no observable behavioral changes up to the point of the
first dynamic type error\footnote{once a dynamic type error is raised,
in languages where the type error can be caught, program behavior may
then further diverge, but this is typically not modeled in gradual
calculi.}.
%
Fundamentally, this property comes down to the behavior of
\emph{runtime type casts}, which implement these generalized runtime
assertions.

@@

\eric{One of the reviewers said this should be moved later in the intro, but I think it makes sense here.}
More concretely, consider a gradually typed language whose only
effects are gradual type errors and divergence. Then if we fix a
result type of natural numbers, a whole program semantics is a partial
function from closed programs to either natural numbers or errors:
\[ -\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N} \cup \{\mho\} \]
where $\mho$ is notation for a runtime type error. We write $M
\Downarrow n$ and $M\Downarrow \mho$ to mean this semantics is defined
as a number or error, and $M\Uparrow$ to mean the semantics is
undefined, representing divergence.
%
A well-behaved semantics should then satisfy several properties. First, it
should be \emph{adequate}: natural number constants should step to
themselves $n \Downarrow n$. Second, it should validate type based reasoning. To
formalize type based reasoning, we give a typed equational theory for
terms of the language $M \equiv N$ for when two terms should be
considered equivalent. Then we want to verify that the big step
semantics respects this equational theory: if closed programs $M \equiv
N$ are equivalent in the equational theory then they have the same
semantics, $M \Downarrow n \iff N \Downarrow n$ and $M\Uparrow \iff N
\Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.
%
Lastly, the graduality property is defined by giving an
\emph{inequational} theory called term precision, where $M \ltdyn N$
roughly means that $M$ and $N$ have the same type erasure and $M$ has
at each point in the program a more precise/static type than $N$.
%
Then, the graduality property states that if $M \ltdyn N$ are whole
programs then $M$ must either have the same behavior as $N$ or error:
Either $M\Downarrow \mho$ or $M \Downarrow n $ and $N \Downarrow n$ or
$M \Uparrow $ and $N \Uparrow$\footnote{we use a slightly more complex
definition of this relation in our technical development below that is
classically equivalent but constructively weaker}.

@@

The main denotational alternative to classical domain theory that can
successfully model these advanced features is \emph{guarded} domain
theory, which we adopt in this work. While classical domain theory is
based on modeling types as ordered sets with certain joins, guarded
domain theory is based on an entirely different foundations, sometimes
(ultra)metric spaces but more commonly as ``step-indexed sets'', i.e.,
objects in the \emph{topos of trees}
\cite{birkedal-mogelberg-schwinghammer-stovring2011}.  Such an object
consists of a family $\{X_n\}_{n \in \mathbb{N}}$ of sets along with
restriction functions $r_n : X_{n+1} \to X_n$ for all $n$.  (in
category theoretic terminology, these are presheaves on the poset of
natural numbers.)  We think a $\mathbb{N}$-indexed set as an infinite
sequence of increasingly precise approximations to the true type being
modeled.
%
Key to guarded domain theory is that there is an operator
$\triangleright$ on step-indexed sets called ``later''. In terms of
sequences of approximations, the later operator delays the
approximation by one step. Then the crucial axiom of guarded domain
theory is that any guarded domain equation $X \cong F(\triangleright
X)$ has a unique solution. This allows guarded domain theory to model
essentially \emph{any} recursive concept, with the caveat that the
recursion is \emph{guarded} by a later.

Since guarded domain theory only provides solutions to guarded domain
equations, there is no systematic way to convert a classical
domain-theoretic semantics to a guarded one.  Classical domain theory
has limitations in what it can model, but it provides \emph{exact}
solutions to domain equations when it applies. When adapting the
New-Licata approach to guarded domain theory, the presence of later in
the semantics makes it \emph{intensional}, unfolding the dynamic type
requires an observable computational step. For example, a function
that pattern matches on an element of the dynamic type and then
returns it unchanged is \emph{not} equal to the identity function,
because it ``costs'' a step to perform the pattern match.
%
This leads to the main departure of our semantics from the New-Licata
approach: while the New-Licata semantics equips $\omega$-CPOs with an
error ordering, we work with types equipped with not just an error
ordering but also a \emph{bisimilarity} relation that reflects when
elements differ only in the number of abstract computational steps
that they take.
%
% ERIC: This could be merged into the summary at the end of the section
The reason for requiring these two separate notions is that in the setting of
guarded domain theory, it is not possible to carry out transitive reasoning
while being oblivious to the number of steps the elements take. We make this
statement precise in \ref{sec:towards-relational-model}. To overcome this
limitation, we develop further changes to the core concepts of our semantics as
discussed in that section.

@@

While guarded domain theory can be presented analytically using
ultrametric spaces or the topos of trees, in practice it is
considerably simpler to work \emph{synthetically} by working in a
non-standard foundational system such as \emph{guarded type
theory}. In guarded type theory later is taken as a primitive
operation on types, and we take as an axiom that guarded domain
equations have a (necessarily unique) solution. The benefit of this
synthetic approach is that when working in the non-standard
foundation, we don't need to model an object language type as a
step-indexed set, but instead simply as a set, and object-language
terms can be modeled in the Kleisli category of a simple monad defined
using guarded recursion. Not only does this make on-paper reasoning
about guarded domain theory easier, it also enables a simpler avenue
to verification in a proof assistant. Whereas formalizing analytic
guarded domain theory would require a significant theory of presheaves
and making sure that all constructions are functors on categories of
presheaves, formalizing synthetic guarded domain theory can be done by
directly adding the later modality and the guarded fixed point
property axiomatically.
%
% \max{TODO: more here, specifically this is where we should talk about adequacy I think}

\max{it's confusing that this is the synthetic section but we switch back to external reasoning. Maybe move this to the previous section and revisit it here?}
Because the solutions we obtain working in guarded domain theory are constructed
as a sequence of increasingly better approximations, when we want to establish a
property that holds in the limit, we must reason about all finite
approximations. In the setting of the topos of trees, this manifests as the need
to consider \emph{global elements} of the presheaves, i.e., a family of elements
$x_i \in X_i$ compatible with the restriction maps. For example, consider the
set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
$i$ times and then returns, and $\bot$ denotes a program that fails to terminate
in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
program that takes $n$ steps and then terminates from a program that never
terminates. On the other hand, if we instead consider the denotation of a
program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
$x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
program will be distinct from that of a program that terminates, regardless of
the number of steps it takes.

\max{shouldn't we say we use clock quantification?}
In the setting of synthetic guarded domain theory, this same idea applies. To
construct a global solution we must use an additional construct, e.g., the
$\square$ modality \cite{10.1007/978-3-319-08918-8_8}, whose semantics in the
topos of trees model is to compute the global elements of a presheaf. A related
approach involves objects known as \emph{clocks}, whereby
\emph{clock-quantification} \cite{atkey-mcbride2013} provides a means to obtain
a global solution.

\max{this next paragraph is a bit too mysterious}
This has important ramifications for defining an adequate denotational semantics
in guarded domain theory, as we seek to do in this paper: we do not want our
denotational semantics to conflate a diverging program with a program that fails
to terminate in sufficiently few steps. Thus, establishing adequacy of a
semantics in guarded domain theory will require a means of globalizing,
a point we will return to in Section \ref{sec:big-step-term-semantics}.

@@

The main contribution of this work is a compositional denotational
semantics for gradually typed languages that validates $\beta\eta$
equality and satisfies a graduality theorem. A great deal of the work
has further been verified in Guarded Cubical Agda \cite{veltri-vezzosi2020}, 
demonstrating that the semantics is readily mechanizable.
We provide an overview of the mechanization effort, and what remains to be formalized,
in Section \ref{sec:mechanization}.

@@

The paper is laid out as follows:
\max{I think there's too much detail in this outline, especially parts 4 and 5. Can we instead expand those into the Intro?}
\begin{enumerate}
\item In Section \ref{sec:GTLC} we fix our input language, a fairly
  typical gradually typed cast calculus.
\item In Section \ref{sec:concrete-term-model} we develop a
  denotational semantics in synthetic guarded domain theory for the
  \emph{terms} of the gradual lambda calculus.  The model is adeqauate
  and validates the equational theory, but it does not satisfy
  graduality. We use this to introduce some of our main technical
  tools: modeling recursive types in guarded type theory and modeling
  effects using call-by-push-value.
\item In Section \ref{sec:towards-relational-model} we show where the
  New-Licata classical domain theoretic approach fails to adapt
  cleanly to the guarded setting and explore the difficulties of
  proving graduality in an intensional model.
  %
  % ... including the reason why we need to introduce weak bisimilarity.
  %
  The key technical issue is that reasoning principles that are simultaneously
  extensional \emph{and} compositional are not compatible with guarded domain
  theory (here, extensional means relating elements that differ only in the
  number of computational steps they take). We make this statement precise with
  a no-go theorem (Theorem \ref{thm:no-go}). In light of this result, to recover
  some amount of compositional reasoning we model the graduality ordering using
  a combination of two relations: a lock-step error ordering in which elements
  must take the same number of steps, and a weak bisimilarity relation where
  elements differing only in their stepping behavior are related.

  % As a result, we split the
  % graduality ordering into two relations: a strong error ordering in which
  % elements must take the same number of steps, and a weak bisimilarity relation
  % where elements differing only in their stepping behavior are related.

\item In Section \ref{sec:concrete-relational-model} we describe the
  construction of the model in detail. Key here is a notion of \emph{syntactic
  perturbations}, which allow us to encode the steps taken by a term in a form
  that we can manipulate. This allows us to recover enough extensional reasoning
  to model the graduality property compositionally.
  % Key here is a notion of \emph{syntactic
  % perturbation}, which give us a type-directed way of imposing delays on terms.
  The resulting model validates the axioms for type and term precision specified
  in Section \ref{sec:GTLC}. The final step is to prove that the model is
  \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
  N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$
  have the same extensional behavior. We do so by extending the globalization
  techniques used in defining the big-step term model to account for the
  interpretation of term precision.
\item In Section \ref{sec:discussion} we discuss prior work on proving
  graduality, the partial mechanization of our results in Agda, and
  future directions for denotational semantics of gradual typing.
\end{enumerate}

>>>>

% \eric{This paragraph could be deleted}
%% In order for this to work as expected, gradually-typed languages should allow for
%% different parts of the codebase to be in different places along the spectrum from
%% dynamic to static, and allow for those different parts to interact with one another.
%% Moreover, gradually-typed languages should support the smooth migration from
%% dynamic typing to static typing, in that the programmer can initially leave off the
%% typing annotations and provide them later without altering the meaning of the program.
%% % Sound gradual typing
%% Furthermore, the parts of the program that are written in a dynamic
%% style should soundly interoperate with the parts that are written in a
%% static style.  That is, the interaction between the static and dynamic
%% components of the codebase should preserve, to the extent possible,
%% the guarantees made by the static types.  In particular, while
%% statically-typed code can error at runtime in a gradually-typed
%% language, such an error can always be traced back to a
%% dynamically-typed term that violated the typing contract imposed by
%% statically typed code. Further, static type assertions are sound in
%% the static portion, and should enable type-based reasoning and
%% optimization.

@@

% Graduality property
One of the fundamental theorems for gradually typed languages is
\emph{graduality}, also known as the \emph{dynamic gradual guarantee},
originally defined by Siek, Vitousek, Cimini, and Boyland
\cite{siek_et_al:LIPIcs:2015:5031, new-ahmed2018}.
%
Informally, graduality says that migrating code from dynamic to
static typing should only allow for the introduction of static or
dynamic type errors, and not otherwise change the behavior of the
program.
%
This is a way to capture programmer intuition that increasing type
precision corresponds to a generalized form of runtime assertions in
that there are no observable behavioral changes up to the point of the
first dynamic type error\footnote{once a dynamic type error is raised,
in languages where the type error can be caught, program behavior may
then further diverge, but this is typically not modeled in gradual
calculi.}.
%
Fundamentally, this property comes down to the behavior of
\emph{runtime type casts}, which implement these generalized runtime
assertions.

@@

% moved the remaining paragraphs to a new section later in the intro

@@

The main denotational alternative to classical domain theory that can
successfully model these advanced features is \emph{guarded} domain
theory, which we adopt in this work. While classical domain theory is
based on modeling types as ordered sets with certain joins, guarded
domain theory is based on an entirely different foundations, sometimes
(ultra)metric spaces but more commonly as ``step-indexed sets'', i.e.,
objects in the \emph{topos of trees}
\cite{birkedal-mogelberg-schwinghammer-stovring2011}.  Such an object
consists of a family $\{X_n\}_{n \in \mathbb{N}}$ of sets along with
restriction functions $r_n : X_{n+1} \to X_n$ for all $n$.  (in
category theoretic terminology, these are presheaves on the poset of
natural numbers.)  We think of a $\mathbb{N}$-indexed set as an infinite
sequence of increasingly precise approximations to the true type being
modeled.
%
Key to guarded domain theory is that there is an operator
$\triangleright$ on step-indexed sets called ``later''. In terms of
sequences of approximations, the later operator delays the
approximation by one step. Then the crucial axiom of guarded domain
theory is that any guarded domain equation $X \cong F(\triangleright
X)$ has a unique solution. This allows guarded domain theory to model
essentially \emph{any} recursive concept, with the caveat that the
recursion is \emph{guarded} by a use of the later operator.

%\max{it's confusing that this is the synthetic section but we switch back to external reasoning. Maybe move this to the previous section and revisit it here?}
Because the solutions we obtain working in guarded domain theory are constructed
as a sequence of increasingly better approximations, when we want to establish a
property that holds in the limit, we must reason about all finite
approximations. In the setting of the topos of trees, this manifests as the need
to consider \emph{global elements} of the presheaves, i.e., a family of elements
$x_i \in X_i$ compatible with the restriction maps. For example, consider the
set of programs that may take a computational step or return unit.\footnote{This example is adapted from \cite{mogelberg-paviotti2016}.} In the
topos of trees, we can model this set as the indexed family of sets $\{X_n\}_{n \in \mathbb{N}}$
where for each $n \ge 1$ we define $X_n = \{0,1,\dots,n-1, \bot\}$. Here, $i$ denotes a program that steps
$i$ times and then returns, and $\bot$ denotes a program that fails to terminate
in $n-1$ steps or fewer. For any fixed $n$, the set $X_n$ fails to distinguish a
program that takes $n$ steps and then terminates from a program that never
terminates. On the other hand, if we instead consider the denotation of a
program to be a global element of the presheaf $\{X_n\}_{n \in \mathbb{N}}$ -- a family of elements
$x_i \in X_i$ for all $i \in \mathbb{N}$ -- then the denotation of the diverging
program will be distinct from that of a program that terminates, regardless of
the number of steps it takes.

% \max{this next paragraph is a bit too mysterious}
% This has important ramifications for defining an adequate denotational semantics
% in guarded domain theory, as we seek to do in this paper: we do not want our
% denotational semantics to conflate a diverging program with a program that fails
% to terminate in sufficiently few steps. Thus, establishing adequacy of a
% semantics in guarded domain theory will require a means of globalizing,
% a point we will return to in Section \ref{sec:big-step-term-semantics}.

While guarded domain theory can be presented analytically using
ultrametric spaces or the topos of trees, in practice it is
considerably simpler to work \emph{synthetically} by working in a
non-standard foundational system such as \emph{guarded type
theory}. In guarded type theory later is taken as a primitive
operation on types, and we take as an axiom that guarded domain
equations have a (necessarily unique) solution. The benefit of this
synthetic approach is that when working in the non-standard
foundation, we don't need to model an object language type as a
step-indexed set, but instead simply as a set, and object-language
terms can be modeled in the Kleisli category of a simple monad defined
using guarded recursion. Not only does this make on-paper reasoning
about guarded domain theory easier, it also enables a simpler avenue
to verification in a proof assistant. Whereas formalizing analytic
guarded domain theory would require a significant theory of presheaves
and making sure that all constructions are functors on categories of
presheaves, formalizing synthetic guarded domain theory can be done by
directly adding the later modality and the guarded fixed point
property axiomatically.
%
% \max{TODO: more here, specifically this is where we should talk about adequacy I think}
%
In this paper, we work informally in a guarded type theory which is described in
more detail in Section \ref{sec:guarded-type-theory}.
%
As in the setting of analytic guarded domain theory, in the synthetic setting it
is also possible to construct global solutions. The specific approach we use in
this work is that of \emph{clocks} and \emph{clock-quantification}
\cite{atkey-mcbride2013}.

% \max{shouldn't we say we use clock quantification?}
% In the setting of synthetic guarded domain theory, there is an analogue of
% constructing global solutions. To construct a global solution we must use an
% additional construct, e.g., the $\square$ modality
% \cite{10.1007/978-3-319-08918-8_8}, whose semantics in the topos of trees model
% is to compute the global elements of a presheaf. A related approach involves
% objects known as \emph{clocks}, whereby \emph{clock-quantification}
% \cite{atkey-mcbride2013} provides a means to obtain a global solution.

\subsection{Adapting the New-Licata Model to the Guarded Setting}

Since guarded domain theory only provides solutions to guarded domain equations,
there is no systematic way to convert a classical domain-theoretic semantics to
a guarded one.  Classical domain theory has limitations in what it can model,
but it provides \emph{exact} solutions to domain equations when it applies. When
adapting the New-Licata approach to guarded domain theory, the presence of later
in the semantics makes it \emph{intensional}: unfolding the dynamic type
requires an observable computational step. For example, a function that pattern
matches on an element of the dynamic type and then returns it unchanged is
\emph{not} equal to the identity function, because it ``costs'' a step to
perform the pattern match.

% - The essential role of transitivity
% - We can't have transitivity + extensional reasoning
% - Solution: split the ordering into two, and introduce perturbations
%   to recover some amount of transitive reasoning

The intensional nature of guarded domain theory leads to the main departure of
our semantics from the New-Licata approach. New and Licata observed that
in order to even \emph{formulate} the model for gradual typing in a
syntax-independent manner, some access to transitive reasoning is essential.
%
However, in the setting of guarded domain theory, it is not possible to carry
out reasoning that is simultaneously transitive and \emph{extensional}, where
the latter means being insensitive to the number of steps the elements take. We
make this statement precise in a no-go theorem (Theorem \ref{thm:no-go}). This fundamental
limitation means that the natural adaptation of the New-Licata semantics to the
guarded setting, where types are interpreted as sets with a
\emph{step-insensitive} error ordering, cannot work, because that ordering is
necessarily \emph{not} transitive.

To recover some amount of compositional reasoning, we model the graduality
ordering using a combination of two relations: a \emph{lock-step error ordering}
in which elements must take the same number of steps, and a \emph{weak
bisimilarity relation} that relates elements differing only in the number of
computational steps that they take. The lock-step ordering \emph{is} transitive,
but it is not sufficient to model the New-Licata proof. The problem is that in
the guarded setting, the denotation of a downcast from the dynamic type involves
a computational step, and by the functorial nature of casts this implies that
the denotations of arbitrary casts may involve computational steps. As a result,
the embedding-projection characterization of casts in the New-Licata model does
not hold in the guarded setting, at least not up to equality.
%
However, while the casts may involve delays, they are otherwise extensionally
equivalent to embedding-projection pairs. Thus, one might try to use weak
bisimilarity to account for the delays, viewing casts as embedding-projection
pairs ``up to weak bisimilarity''. Unfortunately, this does not work: since weak
bisimilarity is \emph{not} transitive (which follows also from our no-go
result), we are back to the original issue of lack of transitive reasoning.

% However, since the casts involve delays but are otherwise extensionally
% equivalent to embedding-projection pairs, one might try to use weak bisimilarity
% to account for the delays. That is, one could try characterizing the casts as
% embedding-projection pairs ``up to weak bisimilarity''. Unfortunately, this does
% not work: since weak bisimilarity is \emph{not} transitive (which follows also
% from our no-go result), we are back to the original issue of lack of
% transitive reasoning.

It would seem that at this point, we are stuck: we need compositional reasoning,
but we also need to be able to account for the steps taken by casts. And now we
come to the key technical contribution of this work. The crucial observation is
that, although the denotations of casts involve delays, the specific ways in
which those delays arise follow a specific, type-directed pattern. Hence, we can
track the delay behavior in a type-directed manner, equipping each type with a
structured set of what we call \emph{syntactic perturbations}. These can then be
used to explicitly \emph{synchronize} elements to ensure that they are in lock-step.
%
In the end, we have come to a resolution of the issue: the lock-step error
ordering where we employ syntactic perturbations to synchronize elements is
still transitive, while simultaneously being able to relate terms that involve
an explicit, known pattern of computational steps.
%
We discuss these issues in more detail in Section
\ref{sec:towards-relational-model} and give a detailed description of the
resulting model in Section \ref{sec:concrete-relational-model}.

% Perturbations allow us to recover enough extensional reasoning to model the
% graduality property compositionally.

% weak bisimilarity: elements differing only in their stepping behavior are related.

% while the New-Licata semantics
% equips $\omega$-CPOs with an error ordering, we work with types equipped with
% not just an error ordering but also a \emph{bisimilarity} relation that reflects
% when elements differ only in the number of abstract computational steps that
% they take. The reason for requiring these two separate notions is that in the
% setting of guarded domain theory, it is not possible to carry out transitive
% reasoning while being oblivious to the number of steps the elements take. We
% make this statement precise in \ref{sec:towards-relational-model}. To overcome
% this limitation, we develop further changes to the core concepts of our
% semantics as discussed in that section.

@@

\subsection{The Graduality Proof}

Proving graduality by constructing a denotational model in guarded type theory
consists of two main steps. First, there is the construction of the model
itself. This gives us an interpretation of the terms as well as the syntactic
axioms that model the graduality property. However, because this model is
defined in the guarded domain theory, while graduality is a statement about a
set-theoretic semantics, it is not a priori clear that the guarded model
validates the graduality property.
%
What we need is to show that the guarded model induces a sensible set-theoretic
semantics in ordinary mathematics. We cannot hope to derive such a semantics for
\emph{all} types (e.g., the dynamic type), but for the subset of closed terms of
base type, we should be able to extract a well-behaved semantics for which the
graduality property holds.

% The next step is to ensure that the semantics actually validates the graduality
% property. That is, we can extract from this guarded model a well-behaved
% semantics in ordinary mathematics For the terms, this means that closed terms of
% base type should evaluate to themselves.

% Defining a model of gradual typing in guarded type theory is not enough to prove
% graduality. We must then extract a well-behaved ``big-step'' semantics for terms
% of base type and show that the denotation of term precision has the expected
% behavior. More specifically, we want a semantics in which the denotation of
% error is different from the denotation of a value, which is in turn different
% from the denotation of the diverging computation. Recall that by working in
% guarded domain theory, our definitions are constructed as a sequence of
% approximations. What we actually want is a statement that holds ``in the
% limit''.

More concretely, consider a gradually typed language whose only effects are
gradual type errors and divergence. If we fix a result type of natural numbers,
a whole program semantics is a partial function from closed programs to either
natural numbers or errors:
%
\[ -\Downarrow : \{M \,|\, \cdot \vdash M : \nat \} \rightharpoonup \mathbb{N}
\cup \{\mho\} \] 
%
where $\mho$ is notation for a runtime type error. We write $M \Downarrow n$ and
$M\Downarrow \mho$ to mean this semantics is defined as a number or error, and
$M\Uparrow$ to mean the semantics is undefined, representing divergence.
%
A well-behaved semantics should then satisfy several properties. First, it
should be \emph{adequate}: natural number constants should evaluate to
themselves: $n \Downarrow n$. Second, it should validate type based reasoning.
To formalize type based reasoning, we give a typed equational theory for terms
of the language $M \equiv N$ for when two terms should be considered equivalent.
Then we want to verify that the big step semantics respects this equational
theory: if closed programs $M \equiv N$ are equivalent in the equational theory
then they have the same semantics, $M \Downarrow n \iff N \Downarrow n$ and
$M\Uparrow \iff N \Uparrow$ and $M \Downarrow \mho \iff N \Downarrow \mho$.

Lastly, the semantics should be adequate for the graduality property. Graduality
is typically axiomatized by giving an \emph{inequational} theory called term
precision, where $M \ltdyn N$ roughly means that $M$ and $N$ have the same type
erasure and $M$ has at each point in the program a more precise/static type than
$N$. Then adequacy says that if $M$ and $N$ are whole programs returning type
$\nat$ and $M \ltdyn N$, then the denotations of $M$ and $N$ as big-step terms
should be related in the expected way: either $M$ errors, or $M$ and $N$ have
the same extensional behavior. That is, either $M\Downarrow \mho$ or $M
\Downarrow n $ and $N \Downarrow n$ or $M \Uparrow $ and $N
\Uparrow$\footnote{we use a slightly more complex definition of this relation in
our technical development below that is classically equivalent but
constructively weaker}.

To define the big-step semantics, we apply clock-quantification to the guarded
model to define a global semantics for closed terms of base type where a term
$M$ is mapped to an element of the coinductive delay monad of Capretta
\cite{lmcs:2265}. We compose this map with a partial function on elements of the
delay monad that returns the result with which given element terminates, or is
undefined if it does not terminate. This gives us the desired partial function
from closed programs to results. We describe this process in more detail in Section
\ref{sec:big-step-term-semantics}.
%
We then make the analogous construction for the relation that denotes the term
precision ordering, showing that a term precision ordering $M \ltdyn N$ implies
the corresponding ordering on the partial functions denoted by $M$ and $N$. This
construction uses as an intermediate step an analogue of the graduality relation
for the coinductive delay monad. The details of the proof are given in Section
\ref{sec:adequacy}.

@@

The main contribution of this work is a compositional denotational semantics in
guarded type theory for a simple gradually typed language that validates
$\beta\eta$ equality and satisfies a graduality theorem.
%
% In particular, the notion of syntactic perturbations stands out as a key
% technical contribution and is the most significant and novel feature of our
% model that allowed us to successfully adapt the denotational approach of New and
% Licata to the guarded setting.
% In particular, the notion of syntactic perturbations stands out as a key feature
% of our model, and is our most significant technical contribution allowing us to
% successfully adapt the denotational approach of New and Licata to the guarded
% setting.
The goal we have kept in mind while designing our model is to handle
gradually-typed languages with advanced features that classical domain theory in
incapable of modeling. We do not address these features in this work but leave
them instead to immediate future work, as handling even the simplest
gradually-typed language already poses interesting challenges.
%
The notion of syntactic perturbation stands out as a key aspect of our model
and our most significant technical contribution, allowing us to successfully
adapt the denotational approach of New and Licata to the guarded setting.
%
Most of our work has further been verified in Guarded Cubical Agda
\cite{veltri-vezzosi2020}, demonstrating that the semantics is readily
mechanizable. We provide an overview of the mechanization effort, and what
remains to be formalized, in Section \ref{sec:mechanization}.
%

% Syntactic perturbations 
%
% The notion of \emph{syntactic perturbations} allow us to encode the steps
% taken by a term in a form that we can manipulate. This allows us to recover
% enough extensional reasoning to model the graduality property compositionally.

% Syntactic perturbations give us a type-directed way of imposing delays on terms.

@@

The paper is laid out as follows:
% \max{I think there's too much detail in this outline, especially parts 4 and 5. Can we instead expand those into the Intro?}
\begin{enumerate}
\item In Section \ref{sec:GTLC} we fix our input language, a fairly
  typical gradually typed cast calculus.
\item In Section \ref{sec:concrete-term-model} we develop a
  denotational semantics in synthetic guarded domain theory for the
  \emph{terms} of the gradual lambda calculus.  The model is adeqauate
  and validates the equational theory, but it does not satisfy
  graduality. We use this to introduce some of our main technical
  tools: modeling recursive types in guarded type theory and modeling
  effects using call-by-push-value.
\item In Section \ref{sec:towards-relational-model} we show where the New-Licata
  classical domain theoretic approach fails to adapt cleanly to the guarded
  setting and explore the difficulties of proving graduality in an intensional
  model. We prove the no-go theorem about extensional, transitive relations,
  introduce the lock-step error ordering and weak bisimilarity relation, and
  motivate the need for perturbations.
\item In Section \ref{sec:concrete-relational-model} we describe the
  construction of the model in detail, and discuss the
  translation of the syntax and axioms of the gradually typed cast calculus into the model.
  Lastly, we prove that the model is adequate for the graduality property.
  % The resulting model validates the axioms for type and term precision specified
  % in Section \ref{sec:GTLC}. The final step is to prove that the model is
  % \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
  % N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$
  % have the same extensional behavior. We do so by extending the globalization
  % techniques used in defining the big-step term model to account for the
  % interpretation of term precision.
\item In Section \ref{sec:discussion} we discuss prior work on proving
  graduality, the partial mechanization of our results in Agda, and
  future directions for denotational semantics of gradual typing.
\end{enumerate}
